---
title: "Storm Data Analysis"
author: "Devender R. Gollapally"
output: 
  html_document: 
    theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Synopsis

Utilizing the storm dataset provided by The U.S National Oceanic and Atmospheric Administration (NOAA) we try to answer some basic questions about severe weather events specifically the following questions:

1. Across the US, which types of events are the most harmful with respect to population health ?
1. Across the US, which types of events have the greatest economic consequences ?

The code used to publish these results are available on GitHub at this [repository](https://github.com/devender/research-project-2).

```{r warning=FALSE,message=FALSE,results="hide",echo=FALSE}
rm(list=ls())
gc()
setwd("/Users/Devender/research-project-2")
library(data.table)
library(dplyr)
library(lubridate)
library(R.utils)
```

## Data Processing
```{r warning=FALSE,message=FALSE,results="hide",echo=FALSE}
# Setup Working Directory
if(!file.exists("data")) { 
    dir.create("data")
}
```
For this research we download the dataset from [here](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2).

```{r warning=FALSE,message=FALSE,results="hide"}
if(!file.exists("data/StormData.csv")){
    
    if(!file.exists("data/StormData.csv.bz2")){ 
        flog.info("downloading dataset")
        url<-"https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
        download.file(url,"data/StormData.csv.bz2",method="curl") 
    }
    
    bunzip2("data/StormData.csv.bz2", destname="data/StormData.csv")
}
```

The unzipped file StormData.csv has more than 30 columns but for this analysis we only require a few, so we read the required data into memory.

```{r warning=FALSE,message=FALSE}
storm_data <- fread("data/StormData.csv",sep=",",header=TRUE,select=c("EVTYPE",
                                                                    "FATALITIES",
                                                                    "INJURIES",
                                                                    "PROPDMG",
                                                                    "PROPDMGEXP",
                                                                    "CROPDMG",
                                                                    "CROPDMGEXP"))

str(storm_data)
```



## Results

